CASE DETECTION OF TUBERCULOSIS PATIENTS USING MACHINE-LEARNING

### 2024-05-05 products/manuscript/Manuscript.qmd
part63

## Introduction
Although early detection and treatment of Tuberculosis cases are the hallmark of successful TB control, diagnostic delay is still long and common in Uganda. Therefore, the study aimed at developing a tool that uses machine-learning techniques to detect and diagnose TB more accurately in a shorter time.

## Methods
This was a retrospective study that used secondary data collected between 2011 to 2018 collected from health under Infectious diseases research collaboration an (IDRC) project called MIND (Mulago Inpatient Non-invasive Diagnosis for pneumonia). The study was conducted on data from 2296 tested patients of whom 1345 had no TB and 951 had TB. Different machine learning metrics(Accuracy, RMSE and AUC) were used evaluate the performance of the models.

## Results
Evaluation of four machine learning models—LASSO, Random Forest, Support Vector Machine (SVM), and Null model—revealed notable performance differences. Notably, the Random Forest model achieved the highest accuracy at 81.57% and an AUC of 0.8868, outperforming the other models.

## Consclusion
The results highlight the potential of advanced machine learning algorithms, particularly Random Forest, in enhancing TB diagnosis and early detection strategies, offering valuable insights for clinical decision-making and the development of more effective screening protocols. {{< pagebreak >}}

# Introduction
Tuberculosis (TB) remains a significant global health challenge due to diagnostic limitations affecting both pediatric and adult populations, resulting in delayed diagnosis or misdiagnosis[@sagili_strategies_2022]. This delay impacts individual prognosis and community transmission. TB, caused by the bacterium Mycobacterium tuberculosis, primarily affects the lungs but can also involve other organs in its active form or remain latent without symptoms[@baykan_extrapulmonary_2022],. In 2020, TB ranked among the top 10 causes of death globally, with over 10 million new cases and 1.3 million deaths reported. Africa bears a considerable burden, with an estimated quarter of new cases globally, resulting in approximately 417,000 deaths annually[@chakaya_who_2022]. TB is a leading cause of death among HIV-infected patients in sub-Saharan Africa, compounded by the low sensitivity of the commonly used sputum smear microscopy, particularly in detecting TB among people living with HIV[@palattiyil_access_2022].

@@

## Questions/Hypotheses to be addressed
The prediction of TB can be modeled using early clinical patient data from Kampala residents who visited the IDRC between 2011 and 2018 and presented with a persistent cough lasting two weeks or more.

{{< pagebreak >}}
# Methods
## Study Design
The study was a retrospective study. This study was based on the secondary data, which was classified as clinical examination, patients' history like alcohol use, smoking and diagnostics.

@@

Random Forest (RF): RF is a versatile and robust ensemble learning method that aims to reduce variance by building multiple decision trees on random subsets of the data and then averaging their predictions. It can handle large datasets with high-dimensional feature spaces and is less prone to overfitting compared to individual decision trees[@simon_interpreting_2023]. However, RF models are more complex and may be challenging to interpret compared to simpler models like decision trees.

@@

Support Vector Machine (SVM): SVM is a powerful supervised learning algorithm commonly used for classification tasks. It works by finding the optimal hyperplane that separates data points belonging to different classes with the maximum margin of separation. SVM is effective in high-dimensional spaces and is versatile in handling linear and non-linear relationships through different kernel functions.[@javatpoint_support_2021] However, SVM can be computationally intensive, especially with large datasets, and may require careful tuning of hyperparameters for optimal performance.

@@

```{r echo=FALSE, message = FALSE }
#| label: tbl-schematic2
#| tbl-cap: "Random Forest"
rf <- readRDS(here("results", "tables", "rfconf_table.rds"))
rf %>%
  knitr::kable()
```

@@

```{r echo=FALSE, message = FALSE}
#| label: tbl-schematic5
#| tbl-cap: "Support Vector Machine"
svm<- readRDS(here("results", "tables", "svmconf_table.rds"))
svm %>%
  knitr::kable()
```

@@

```{r echo=FALSE, message = FALSE}
#| label: tbl-schematic3
#| tbl-cap: "lassoconf_table"
ls<- readRDS(here("results", "tables", "lassoconf_table.rds"))
ls %>%
  knitr::kable()
```

@@

# Models Evaluation

@@

The study presents a comprehensive analysis of machine learning models' efficacy in predicting the presence of tuberculosis (TB) based on a dataset comprising 2296 instances and 14 variables associated with TB. The study selected nine variables for prediction after conducting chi-square tests to eliminate non-statistically significant variables associated with the outcome. The retained variables for prediction included asthma, alcohol use, fever, chest radiography, home fuel use, weight loss, presence of blood in sputum, oxygen saturation, and HIV status. The study evaluated four distinct machine learning models, namely LASSO, Random Forest, Support Vector Machine (SVM), and a Null model, for their predictive performance using key metrics accuracy and Area Under the Curve (AUC). The LASSO model achieved an accuracy of 80.26% and an AUC of 0.8895. The Random Forest model outperformed others with an accuracy of 81.57% and an AUC of 0.8868. SVM achieved an accuracy of 79.68% and an AUC of 0.8825. The Null model achieved an accuracy of 60.23% and an AUC of 0.5.

Furthermore, the study presented detailed classification results for each model, emphasizing their ability to correctly classify true positives and true negatives while minimizing false positives and false negatives. Particularly, the Random Forest model demonstrated robust performance in correctly identifying true positives and true negatives, with minimal misclassifications. The findings underscore the potential of advanced machine learning algorithms, particularly Random Forest and LASSO, in enhancing TB diagnosis and early detection strategies. The manuscript provides valuable insights into the predictive capabilities of different machine learning models which can inform clinical decision-making and aid in the development of more effective TB screening tolls and protocols.

# Strengths

In this study, a rigorous variable selection process was implemented, utilizing chi-square tests to identify statistically significant predictors associated with tuberculosis (TB) from a dataset comprising 2296 instances and 14 variables. By employing stringent selection criteria, the study ensures that only variables with meaningful associations with TB are included in subsequent analyses, thereby enhancing the model's interpretability, and reducing the risk of overfitting. Moreover, the evaluation of multiple machine learning models, including LASSO, Random Forest, Support Vector Machine (SVM), and a Null model, contributes to a comprehensive comparison of different predictive approaches. This assessment allows for informed decision-making regarding model selection, facilitating the identification of the most suitable algorithm for TB prediction.

Additionally, the study employs robust performance metrics such as accuracy and Area Under the Curve (AUC) to evaluate model performance objectively. These metrics provide a comprehensive assessment of each model's predictive capabilities, enabling comparisons across different algorithms. Furthermore, the detailed classification results presented offer insights into the models' ability to correctly classify true positives and true negatives while minimizing false positives and false negatives.

# Limitations and Recommendations

Despite the strengths, this study had limitations that warrant consideration. Firstly, the dataset's potential class imbalance, characterized by variations in the number of instances for individuals with and without TB, may impact model performance of the model. Additionally, while the variable selection methodology employed chi-square tests, alternative feature selection methods like recursive feature elimination could offer further insights into the predictive variables' relevance and importance. Future research should aim incorporate larger, more diverse datasets and conducting external validation studies to assess model performance in real-world clinical settings and use of more robust techniques such as neural networks.

# Conclusions

>>>>

## Introduction
Although early detection and treatment of Tuberculosis cases are the hallmarks of successful TB control, diagnostic delays are still long and common in Uganda. Therefore, this study aimed to develop a tool that utilizes machine-learning techniques to detect and diagnose TB more accurately within a shorter time frame.

## Methods
This was a retrospective study that utilized secondary data collected between 2011 and 2018 from health facilities under the Infectious Diseases Research Collaboration (IDRC) project called MIND (Mulago Inpatient Non-invasive Diagnosis for pneumonia). The study analyzed data from 2296 tested patients, among whom 1345 were found to have no TB and 951 were diagnosed with TB. Various machine learning metrics (Accuracy, RMSE, and AUC) were used to evaluate the performance of the models.

## Results
Evaluation of four machine learning models, LASSO, Random Forest, Support Vector Machine (SVM), and Null model—revealed notable performance differences. Particularly, the Random Forest model achieved the highest accuracy at 81.57% and an AUC of 0.8868, outperforming the other models.

## Conclusion
The results underscore the potential of advanced machine learning algorithms, particularly Random Forest, in enhancing TB diagnosis and early detection strategies. These findings offer valuable insights for clinical decision-making and the development of more effective screening protocols.

# Introduction
Tuberculosis (TB) remains a significant global health challenge due to diagnostic limitations affecting both pediatric and adult populations, resulting in delayed diagnosis or misdiagnosis[@sagili_strategies_2022]. This delay impacts individual prognosis and community transmission. TB caused by the bacterium Mycobacterium tuberculosis primarily affects the lungs but can also involve other organs in its active form or remain latent without symptoms[@baykan_extrapulmonary_2022]. In 2020, TB ranked among the top 10 causes of death globally, with over 10 million new cases and 1.3 million deaths reported. Africa bears a considerable burden with an estimated quarter of new cases globally, resulting in approximately 417,000 deaths annually[@chakaya_who_2022]. TB is a leading cause of death among HIV-infected patients in sub-Saharan Africa, compounded by the low sensitivity of the commonly used sputum smear microscopy, particularly in detecting TB among people living with HIV[@palattiyil_access_2022].

@@

## Questions/Hypotheses to be addressed
How can the prediction of TB be modeled  early using clinical patient data from residents of Kampala who visited the IDRC between 2011 and 2018 who had cough for more than two weeks?

# Methods
## Study Design
The study was a retrospective study. This study was based on the secondary data, which was classified as clinical examination, patients' history like alcohol use, smoking and diagnostics.

@@

Random Forest (RF): RF is a versatile and robust ensemble learning method that aims to reduce variance by building multiple decision trees on random subsets of the data and then averaging their predictions. It can handle large datasets with high-dimensional feature spaces and is less prone to over fitting compared to individual decision trees[@simon_interpreting_2023]. However, RF models are more complex and may be challenging to interpret compared to simpler models like decision trees.

@@

Support Vector Machine (SVM): SVM is a powerful supervised learning algorithm commonly used for classification tasks. It works by finding the optimal hyperplane that separates data points belonging to different classes with the maximum margin of separation. SVM is effective in high dimensional spaces and is versatile in handling linear and non-linear relationships through different kernel functions.[@javatpoint_support_2021] However, SVM can be computationally intensive, especially with large datasets, and may require careful tuning of hyperparameters for optimal performance.

@@

```{r echo=FALSE, message = FALSE }
#| label: tbl-schematic2
#| tbl-cap: "Random Forest Predictions"
rf <- readRDS(here("results", "tables", "rfconf_table.rds"))
rf %>%
  knitr::kable()
```

@@

```{r echo=FALSE, message = FALSE}
#| label: tbl-schematic5
#| tbl-cap: "Support Vector Machine Predictions"
svm<- readRDS(here("results", "tables", "svmconf_table.rds"))
svm %>%
  knitr::kable()
```

@@

```{r echo=FALSE, message = FALSE}
#| label: tbl-schematic3
#| tbl-cap: "LASSO Predictions"
ls<- readRDS(here("results", "tables", "lassoconf_table.rds"))
ls %>%
  knitr::kable()
```

@@

## Models Evaluation

@@

The study offers a comprehensive analysis of machine learning models' effectiveness in predicting tuberculosis (TB) presence using a dataset containing 2296 instances and 14 variables associated with TB. The study employed cross-validation with the LASSO method to identify significant predictors of the outcome. Three distinct machine learning models, LASSO, Random Forest, and Support Vector Machine (SVM) were evaluated for predictive performance using accuracy and Area Under the Curve (AUC) metrics.
The LASSO model achieved an accuracy of 80.26% and an AUC of 0.8895, while the Random Forest model outperformed others with an accuracy of 81.57% and an AUC of 0.8868. SVM achieved an accuracy of 79.68% and an AUC of 0.8825. In contrast, the Null model achieved an accuracy of 60.23% and an AUC of 0.5. Detailed classification results were presented for each model, highlighting their ability to correctly classify true positives and true negatives while minimizing false positives and false negatives. Notably, the Random Forest model demonstrated robust performance in accurately identifying true positives and true negatives with minimal misclassifications.

## Strengths

Employing cross-validation helped enhance the model’s interpretability and reduce the risk of overfitting. Moreover, evaluating multiple machine learning models, including LASSO, Random Forest, Support Vector Machine (SVM), and a Null model, contributes to a comprehensive comparison of different predictive approaches. This assessment allows for informed decision-making regarding model selection, facilitating the identification of the most suitable algorithm for TB prediction. Additionally, the study employs robust performance metrics such as accuracy and Area Under the Curve (AUC) to evaluate model performance objectively. These metrics provide a comprehensive assessment of each model’s predictive capabilities, enabling comparisons across different algorithms. Furthermore, the detailed classification results presented offer insights into the models’ ability to correctly classify true positives and true negatives while minimizing false positives and false negatives.
Additionally, the study employs robust performance metrics such as accuracy and Area Under the Curve (AUC) to evaluate model performance objectively. These metrics provide a comprehensive assessment of each model's predictive capabilities, enabling comparisons across different algorithms. Furthermore, the detailed classification results presented offer insights into the models' ability to correctly classify true positives and true negatives while minimizing false positives and false negatives.

## Limitations and Recommendations

Despite the strengths, this study had limitations that warrant consideration. Firstly, the dataset's potential class imbalance, characterized by variations in the number of instances for individuals with and without TB, may impact model performance of the model. Future research should aim incorporate larger, more diverse datasets and conducting external validation studies to assess model performance in real world clinical settings and use of more robust techniques such as neural networks.

## Conclusions
